Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-09 write-ups/PLDI/paper.tex

\section{The ckill and cfork primitives}

>>>>

\section{The Cooperative Kernels Programming Model}

We describe our cooperative kernels programming model as an extension
to the industry-standard OpenCL programming model for heterogenous
many-core architectures.  However, the concept behind our programming
model is more general, and could be applied to extend other GPU
programming models, such as CUDA and HSA.

\subsection{Semantics of Cooperative Kernels}

As with a regular OpenCL kernel, a cooperative kernel is launched by
the host application, with a specified number of workgroups, each
consisting of a specified number of threads.  The kernel's behaviour
is described by a kernel function, written in OpenCL C but using the
additional language extensions that we present below, and a number of
parameters to the kernel function are passed by the host application.

\paragraph{Restored variables}

\ADComment{TODO; perhaps say that kernel parameters are immutable.}

\paragraph{Initial workroups}

If the host application launches a cooperative kernel specifying that
$N$ workgroups should execute the kernel, then when the kernel
commences execution it should be executed by some number of workgroups
$M$, with $1\leq M \leq N$.  The requested number of workgroups is
called the \emph{workgroup limit} for the cooperative kernel, and $M$
is the number of \emph{active workgroups}.

The initial workgroups have consecutive workgroup ids in the range
$[0, M-1]$.

\paragraph{Semantics for $\offerkill$}

Our first additional language primitive, $\offerkill$, is a
workgroup-level function.  That is, similar to the OpenCL
intra-workgroup $\mathsf{barrier}$ primitive, if $\offerkill$ appears
inside conditional code then it must be executed by all threads in a
workgroup, with all threads having executed the same number of
iterations of any enclosing loops.

Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
i.e.\ the workgroup's id is not the largest among the active
workgroups, then $\offerkill$ is a no-op.  Otherwise, we must have $m
= M-1$, in which case there is a non-deterministic choice between (a)
$\offerkill$ executing as a no-op, or (b) execution of the workgroup
ceasing and the number of active workgroups $M$ being decremented by
one.  Case (b) corresponds to the scheduler accepting the workgroup's
offer to cease execution, allowing the scheduler to schedule some
other task.

The semantics of $\offerkill$ is atomic.  That is, if multiple
workgroups reach $\offerkill$ simultaneously, the semantics is that
the workgroups should each execute $\offerkill$ in some well-defined
but arbitrary sequential order.  \ADComment{Should we say something
  about linearisability?}

\paragraph{Semantics for $\offerfork$}

Our second language primitive, $\offerfork$, is also a workgroup-level
function.

Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
$k$ new workgroups are spawned with consecutive ids in the range $[M,
  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.
The $k$ new workgroups commence execution at the program point
immediately following the $\offerfork$ call.  Each thread in each new
workgroup has an arbitrary local state, except that:

\begin{itemize}

\item the threads have access to the parameters originally passed to
  the cooperative kernel by the host application (recall that these
  are immutable);

\item for each variable $v$ in the set of restored variables
  identified for the cooperative kernel, every thread's copy of $v$ is
  initialised to the value that thread 0 in workgroup $m$ held for $v$
  at the point of the $\offerfork$ call.

\end{itemize}

Notice that $k=0$ is always a legitimate choice for the number of
workgroups to be spawned by an $\offerfork$ call, and in particular
that if the number of active workgrouups $M$ is equal to the workgroup
limit $N$, $k=0$ is guaranteed.

\ADComment{Remark on our decision to have every thread inherit the state of the master thread from workgroup $m$.}

\subsection{Non-functional Requirements of Cooperative Kernels}

If a cooperative kernel is launched with $M$ workgroups, the

However, unlike a regular kernel, if $M$ workgroups are requested for execution of a cooperative kernel.

A \emph{cooperative kernel} is launched with a desired workgroup
count, $M$, where $M > 0$.
